---
title: "Simulating for understanding p-values"
author: "Jean-Jacques Orban de Xivry"
date: "April 8, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction



When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## simulating a single correlation R=0

If we draw random numbers from two different normal distributions with mean=0 and standard deviation=1, we know that the correlation between these two random variables should be non-significant as there is not link between them. In R, we can simulate such

## What is the expected p-value distribution if there are no correlations

## simulating a single correlation R=x

## What is the expected p-value distribution if there is a correlation

## simulating multiple correlations, bootstrap

## How sensisitive are correlations to one outlier?

## Solution to the outlier problem
### robust parametric correlation
### non-parametric correlation - Spearman

## How sensisitive are correlations to subgroups?

## Solution to the subgroup problem
### Spearmann is not a solution
### Taking subgroups into account

## The multiple comparison problem

## extending this line of thought to the T-test



You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
